---
title: "geofi R-package for accessing Statistics Finland spatial data"
author: "Markus Kainu"
date: 2020-02-11 10:53:45
categories: 
  - research
  - data
slug: geofi-en
tags:
  - R-bloggers
  - R
  - graphics
  - dataviz
  - ropengov
draft: false
editor_options: 
  chunk_output_type: console
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      cache = FALSE,
                      fig.width=12,
                      fig.height=10,
                      dpi = 60, 
                      class.output = "img-fluid",
                      out.width=900)

# Oletusteema
library(ggplot2)
theme_set(
  theme_minimal(base_family = "PT Sans", base_size = 12)
)
options(scipen = 999)
```

We have slowly started developing a follow-up package with [ropengov](https://ropengov.github.io/)-posse for [gisfin](https://ropengov.github.io/gisfin/)-package named [`geofi`](https://ropengov.github.io/geofi/index.html). Package provides access to few sources of *Finnish open geospatial data* from R. We are focusing in administrative regions at the moment and our primary source of data is Statistics Finland and their `wfs`-api. You can use functions in  `geofi` fecth data such as *municipality borders*, *postal code areas* sekä *population grids*. 


`geofi` is not published in CRAN yet and you cant install it using `intall.packages()`. But you can install it directly from Github with `remotes::install_github("ropengov/geofi")` and try out the following examples.  For quick access try our Shiny app at:  [`https://muuankarski.shinyapps.io/geofi_selain/`](https://muuankarski.shinyapps.io/geofi_selain/).


**Municipalility borders**

```{r municipality_map}
library(geofi)
library(ggplot2)

municipalities <- get_municipalities(year = 2020, scale = 4500)
ggplot(municipalities) + 
  geom_sf(aes(fill = as.integer(kunta))) +
  scale_fill_viridis_c()
```


**Postal code areas**

```{r zipcode_map}
zipcodes <- get_zipcodes(year = 2020) 
ggplot(zipcodes) + 
  geom_sf(aes(fill = as.integer(posti_alue)), color = alpha("white", 1/3)) +
  scale_fill_viridis_c()
``` 

**Population grids**

```{r, population_grid_data}
pop_grid <- get_population_grid(year = 2018, resolution = 5)
ggplot(pop_grid) + 
  geom_sf(aes(fill = objectid), color = alpha("white", 1/3)) +
  scale_fill_viridis_c()
```


Regions (maakunnant), health care districts (sairaanhoitopiirit) and many more regional breakdowns are based on municipality divide.  `get_municipalities()`-function returns data containing [these attribute variables](https://ropengov.github.io/geofi/reference/municipality_key_2020.html) (year 2020), that you can use to aggregate from municipality level upwards.


```{r, aggregate}
library(dplyr)
municipalities <- get_municipalities(year = 2019, scale = 4500)
regions <- municipalities %>% 
  group_by(maakunta_name_fi) %>% summarise()
ggplot(regions) + 
  geom_sf(aes(fill = maakunta_name_fi)) +
  scale_fill_viridis_d()
```

You can join `geofi` datas with other attribute, too.  Below is an example on how to get (non-spatial statistical) data from Statistics Finland and create a map on population at municipality level.


```{r municipalities_with_data}
library(tidyr)
library(pxweb)
library(janitor)

municipalities17 <- get_municipalities(year = 2017)

# pull municipality data from Statistics Finland
pxweb_query_list <-
  list("Alue 2019"=c("*"),
       "Tiedot"=c("*"),
       "Vuosi"=c("2017"))
px_data <-
  pxweb_get(url = "http://pxnet2.stat.fi/PXWeb/api/v1/fi/Kuntien_avainluvut/2019/kuntien_avainluvut_2019_aikasarja.px",
            query = pxweb_query_list)
# Convert to data.frame
tk_data <- as.data.frame(px_data, column.name.type = "text", variable.value.type = "text")
tk_data2 <- tk_data %>%
  rename(name = `Alue 2019`) %>%
  mutate(name = as.character(name),
         # Paste Tiedot and Vuosi
         Tiedot = paste(Tiedot, Vuosi)) %>%
  select(-Vuosi) %>%
  spread(Tiedot, `Kuntien avainluvut`) %>%
  as_tibble()
tk_data3 <- janitor::clean_names(tk_data2)

# Join with Statistics Finland attribute data
dat <- left_join(municipalities17, tk_data3)

ggplot(dat) + 
  geom_sf(aes(fill = vakiluku_2017), color = alpha("white", 1/3)) +
  scale_fill_viridis_c(trans = "sqrt")
```

Take a look at the Github-site and join us!
